---
title: "Peer-graded Assignment: Reproducible Research Course Project 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
### Peer-graded Assignment: Reproducible Research Course Project 1
### Objective: Collect, work with, and analyze the given dataset
### Astudent44
---


## Loading and preprocessing the data
#### Load in libraries
```{r}
library(ggplot2)
library(plyr)
library(lubridate)
```

#### Assume that raw data file is in the working directory
```{r}
dataset <- read.csv("activity.csv")
```

#### Clean data by formatting date and removing na
```{r}
dataset$date <- ymd(dataset$date)
cleandata <- dataset[!is.na(dataset$steps),]
```



## What is mean total number of steps taken per day?
#### Create a summary table of dates and steps
```{r}
stepsPerDay <- aggregate(cleandata$steps ~ cleandata$date, FUN = sum,)
colnames(stepsPerDay) <- c("Date", "Steps")
```

#### Graph steps per day, total steps by frequency
```{r}
png("plot1.png", height = 480, width = 480)
hist(as.numeric(stepsPerDay$Steps), 
     main = "Total Number of Steps Taken Each Day", 
     xlab = "Steps", col = "Blue")
dev.off()
```

```{r echo = FALSE}
hist(as.numeric(stepsPerDay$Steps), 
     main = "Total Number of Steps Taken Each Day", 
     xlab = "Steps", col = "Blue")
```

#### Calculate mean and median steps taken per day
```{r}
print(paste("Mean Number of Steps Taken Each Day = ",
            mean(stepsPerDay$Steps)))
print(paste("Median Number of Steps Taken Each Day = ",
            median(stepsPerDay$Steps)))
```



## What is the average daily activity pattern?
#### Make a time series plot of the 5-minute interval and the average steps
```{r}
avgDailyActivity <- aggregate(cleandata$steps, 
                              by = list(cleandata$interval), FUN = mean)
colnames(avgDailyActivity) <- c("Interval", "Mean")
```

#### Plot average daily activity as a line graph
```{r}
png("plot2.png", height = 480, width = 640)
plot(avgDailyActivity$Interval, avgDailyActivity$Mean, type = "l", 
     col = "Blue", lwd = 2, xlab = "Interval (by 5-minute)", 
     ylab = "Average Number of Steps Taken Averaged Across All Days",
     main = "Average Daily Activity Pattern")
dev.off()
```

```{r echo = FALSE}
plot(avgDailyActivity$Interval, avgDailyActivity$Mean, type = "l", 
     col = "Blue", lwd = 2, xlab = "Interval (by 5-minute)", 
     ylab = "Average Number of Steps Taken Averaged Across All Days",
     main = "Average Daily Activity Pattern")
```

#### Determine which interval has the greatest number of steps
```{r}
intervalSteps <- aggregate(cleandata$steps ~ cleandata$interval, FUN = sum,)
colnames(intervalSteps) <- c("Interval", "Steps")
print(paste("The 5-minute with the Maximum Number of Steps = ",
            intervalSteps[which.max(intervalSteps$Steps),]$Interval))
```



## Imputing missing values
#### Calculate the number of NA values 
```{r}
print(paste("The Total Number of Missing Values in the Dataset = ",
            sum(is.na(dataset$steps))))
```

#### Create new data set with NA values filled in
#### Strategy used is to fill in with mean for that day
```{r}
avgSteps <- aggregate(dataset$steps ~ dataset$date, FUN = mean,)
colnames(avgSteps) <- c("Date", "Steps")
filleddata <- transform(dataset, steps = ifelse(is.na(dataset$steps), 
                                                yes = avgSteps$Steps,
                                                no = dataset$steps))
```

#### Redo part 1 with the new data set
#### Graph steps per day, total steps by frequency
#### Calculate mean and median steps taken per day
```{r}
filledstepsPerDay <- aggregate(filleddata$steps ~ filleddata$date, FUN = sum,)
colnames(filledstepsPerDay) <- c("Date", "Steps")
png("plot3.png", height = 480, width = 480)
hist(as.numeric(filledstepsPerDay$Steps), 
     main = "Filled Total Number of Steps Taken Each Day", 
     xlab = "Steps", col = "Blue")
dev.off()
print(paste("Filled Mean Number of Steps Taken Each Day = ",
            mean(filledstepsPerDay$Steps)))
print(paste("Filled Median Number of Steps Taken Each Day = ",
            median(filledstepsPerDay$Steps)))
print(paste("These values DO differ from those of part 1.  By replacing",
            " the NA values with the average number of steps taken",
            " per day the mean and median increase because the average",
            " is greater than the lowest number of steps per day."))
```

```{r echo = FALSE}
hist(as.numeric(filledstepsPerDay$Steps), 
     main = "Filled Total Number of Steps Taken Each Day", 
     xlab = "Steps", col = "Blue")
```


## Are there differences in activity patterns between weekdays and weekends?
#### Add new variable column to the filled data set
```{r}
dayName <- weekdays(filleddata$date)
filledweekdata <- cbind(filleddata, dayName)
filledweekdata$dateType <- ifelse(filledweekdata$dayName %in% c("Saturday", "Sunday"),
                                  "Weekend", "Weekday")
```

#### Plot average daily activity as a line graph for dateType
```{r}
avgDayDailyActivity <- aggregate(steps~interval + dateType, 
                                 filledweekdata, FUN = mean)
png("plot4.png", height = 480, width = 640)
plot<- ggplot(avgDayDailyActivity, aes(x = interval , y = steps, color = dateType)) +
  geom_line() +
  labs(title = "Average Daily Activity Pattern by Date Type", 
       x = "Interval (by 5-minute)", 
       y = "Average Number of Steps") +
  facet_wrap(~dateType, ncol = 1, nrow=2)
print(plot)
dev.off()
```

```{r echo = FALSE}
plot<- ggplot(avgDayDailyActivity, aes(x = interval , y = steps, color = dateType)) +
  geom_line() +
  labs(title = "Average Daily Activity Pattern by Date Type", 
       x = "Interval (by 5-minute)", 
       y = "Average Number of Steps") +
  facet_wrap(~dateType, ncol = 1, nrow=2)
print(plot)
```
